---
title: "Report Template"
author: "YOUR NAME (your NetID)"
date: "DATE"
format: pdf
fontsize: 12pt
number-sections: TRUE
linkcolor: purple
geometry:
  - margin=1in
toc: TRUE
---

# Abstract {#sec-abstract}

\begin{abstract}
When writing an academic-style paper, it is important to include an abstract. The abstract is a short, few-line summary of the paper, and should include major findings and conclusions.
\end{abstract}


# Section Header {#sec-section1}

Notice how the section headers are encoded as first-level headers (using a single hashtag \#). It is recommended to use first-level headers for main section headers, second-level headers for subsection headers, etc.

## Subsection 1

For example, we could use this subsection to discuss something specific about the overall topic of @sec-section1. By the way, notice how we managed to include a link to a previous section, using the section tag we created earlier (in the curly braces \{ \}).

# Equations

Equations follow standard LaTeX syntax, like
$$ \mathbb{E}[X] := \int_{-\infty}^{\infty} x f_X(x) \ \mathrm{d}x $$
or $f(x) = x^2$.

# Table of Contents

If you would like to remove the table of contents at the start of your report, simply change the value of `toc:` in the YAML header from `TRUE` to `FALSE`. 

# Code

One of the major benefits of using Quarto to generate reports is that we can easily integrate code. For instance:
```{r}
x <- 2
x + 2
```

Packages like `pander` can help render tabular output in slightly nicer formats. For instance, consider the data frame

```{r}
my_df <- data.frame(
  col1 = c(1, 2, 3, 4),
  col2 = c("a", "b", "c", "d")
)

my_df
```

Using `pander`, we could instead have

```{r}
library(pander)

pander(my_df)
```

A similar package is the `kableExtra` package; feel free to look up the help file on this package, and use it instead of `pander` if you like!